\textbf{DiracDec C++/Coq Implementation\cite{ExampleCitation}}

### 2024-12-14 main.tex
de Bruijn expression rules updated

\section{TODO}
\begin{itemize}
    \item Alpha equivalence is checked by a bound variable renaming on the whole term.
    \item arrow type of $\Type \to \Type$ are forbidden. Check whether typing rules comply with this restriction.
\end{itemize}

@@

Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_{i \in s} X$ to denote $\sum_{s} \lambda i : T . X$.

@@

\clearpage
\section{Conversions and Reductions}
\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda : T.t)\ u)\ \reduce\ t[u]} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda.t)\ u)\ \reduce\ t[u]} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
        \\
        \textbf{Eta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ x)\ \reduce\ t}
        \qquad \qquad
        \textbf{Eta-Index} \qquad
        \frac{}{E[\Gamma] \vdash \lambda . (t\ x)\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

\begin{ruletable}{Rules for sum operators.}
    %
    R-SUM-CONST0
    & $ \sum_{x \in s} 0 \ \reduce\ 0 $ \\
    %
    R-SUM-CONST1
    & $ \sum_{x \in s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
    %
    R-SUM-CONST2
    & $ \sum_{x \in s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
    %
    R-SUM-CONST3
    & $ \sum_{x \in s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
    %
    R-SUM-CONST4
    & $ \ONEO(\sigma) \ \reduce\ \sum_{i \in \mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
\end{ruletable}

\begin{ruletable}{Rules for eliminating $\delta_{s, t}$. Note that these rules will match the $\delta$ operator modulo the commutativity of its arguments.}
    %
    R-SUM-ELIM0
    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} \delta_{i, t} \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  1$ \\
    %
    R-SUM-ELIM1
    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n) $ \\
    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} a_1\{i/t\} \times \cdots \times a_n\{i/t\} $ \\
    %
    R-SUM-ELIM2
    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, t}.A) \ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{i/t\} $ \\
    %
    R-SUM-ELIM3
    & $ i \text{ free in } t \vdash \sum_{i \in \mathbf{U}(\sigma)} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, t} \times \cdots \times a_n).A $ \\
    & $ \reduce\ \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  (a_1\{i/t\} \times \cdots \times a_n\{i/t\}).A\{i/t\} $ \\
    R-SUM-ELIM4
    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n}  \delta_{i, j} \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} 1 $ \\
    %
    R-SUM-ELIM5
    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \dots \times \delta_{i, j} \times \cdots \times a_n) $ \\
    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}) $ \\
    %
    R-SUM-ELIM6
    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (\delta_{i, j}.A) \ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} A\{j/i\} $ \\
    %
    R-SUM-ELIM7
    & $ \sum_{i \in M} \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1 \times \cdots \times \delta_{i, j} \times \cdots \times a_n).A $ \\
    & $ \reduce\ \sum_{j \in M} \sum_{k_1 \in s_1} \cdots \sum_{k_n \in s_n} (a_1\{j/i\} \times \cdots \times a_n\{j/i\}).A\{j/i\} $ \\
\end{ruletable}

\begin{ruletable}{Rules for pushing terms into sum operators}
    %
    R-SUM-PUSH0
    & $ b_1 \times \cdots \times (\sum_{i \in M} a) \times \cdots \times b_n \ \reduce\ \sum_{i \in M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
    %
    R-SUM-PUSH1
    & $ (\sum_{i \in M} a)^* \ \reduce\ \sum_{i \in M} a^* $ \\
    %
    R-SUM-PUSH2
    & $ (\sum_{i \in M} X)^\dagger \ \reduce\ \sum_{i \in M} X^\dagger $ \\
    %
    R-SUM-PUSH3
    & $ a.(\sum_{i \in M} X) \ \reduce\ \sum_{i \in M} (a.X) $ \\
    %
    R-SUM-PUSH4
    & $ (\sum_{i \in M} a).X \ \reduce\ \sum_{i \in M} (a.X) $ \\
    %
    R-SUM-PUSH5
    & $ (\sum_{i \in M} B)\cdot K \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
    %
    R-SUM-PUSH6
    & $ (\sum_{i \in M} O)\cdot K \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
    %
    R-SUM-PUSH7
    & $ (\sum_{i \in M} B)\cdot O \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
    %
    R-SUM-PUSH8
    & $ (\sum_{i \in M} K)\cdot B \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
    %
    R-SUM-PUSH9
    & $ (\sum_{i \in M} O_1)\cdot O_2 \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
    %
    R-SUM-PUSH10
    & $ B \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(B \cdot K) $ \\
    %
    R-SUM-PUSH11
    & $ O \cdot (\sum_{i \in M} K) \ \reduce\ \sum_{i \in M}(O \cdot K) $ \\
    %
    R-SUM-PUSH12
    & $ B \cdot (\sum_{i \in M} O) \ \reduce\ \sum_{i \in M}(B \cdot O) $ \\
    %
    R-SUM-PUSH13
    & $ K \cdot (\sum_{i \in M} B) \ \reduce\ \sum_{i \in M}(K \cdot B) $ \\
    %
    R-SUM-PUSH14
    & $ O_1 \cdot (\sum_{i \in M} O_2) \ \reduce\ \sum_{i \in M}(O_1 \cdot O_2) $ \\
    %
    R-SUM-PUSH15
    & $ (\sum_{i \in M} X_1) \otimes X_2 \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $ \\
    %
    R-SUM-PUSH16
    & $ X_1 \otimes (\sum_{i \in M} X_2) \ \reduce\ \sum_{i \in M} (X_1 \otimes X_2) $
\end{ruletable}
\yx{Note: because we apply type checking on variables, and stick to unique bound variables, these pushing in operations are always sound.}

\begin{ruletable}{Rules for addition and index in sum}
    %
    R-SUM-ADDS0
    & $\sum_{i \in M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{i \in M} a_1) + \cdots + (\sum_{i \in M} a_n) $ \\
    %
    R-SUM-ADD0
    & $\sum_{i \in M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{i \in M} X_1) + \cdots + (\sum_{i \in M} X_n) $ \\
    %
    R-SUM-ADD1
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
    %
    R-SUM-ADD2
    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
    %
    R-SUM-ADD3
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
    %
    R-SUM-INDEX0
    & $ \sum_{i \in \mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{j \in \mathbf{U}(\sigma)} \sum_{k \in \mathbf{U}(\tau)} A\{i/(j, k)\} $ \\
    %
    R-SUM-INDEX1
    & $ \sum_{i \in M_1 \star M_2} A \ \reduce\ \sum_{j \in M_1} \sum_{k \in M_2} A\{i/(j, k)\} $ \\
    %
    R-SUM-SWAP
    & $ M_1 < M_2 \vdash \sum_{i \in M_2} \sum_{j \in M_1} X \ \reduce\ \sum_{j \in M_1} \sum_{i \in M_2} X $
\end{ruletable}

>>>>

\section{TODO}
\begin{itemize}
    \item \ 
\end{itemize}

\section{Some practices of de Bruijn substitution and application}
\begin{itemize}
    \item $\lambda x .((\lambda y. \lambda z. \lambda w. y)\ x) \reduce \lambda x. \lambda z. \lambda w. x $ 
    \\
    $ \lambda . ((\lambda . \lambda . \lambda. 2)\ 0) \reduce \lambda . \lambda . \lambda.2$

    \item $\lambda x . ((\lambda y. \lambda z. y)\ (X\ x)) \reduce \lambda x . \lambda z . (X\ x)$
    \\
    $\lambda . ((\lambda. \lambda. 1)\ (X\ 0)) \reduce \lambda. \lambda. (X\ 1)$

\end{itemize}

@@

Usually, the sum body is specified by an abstraction. Therefore we use notation $\sum_s X$ to denote $\sum_{s} \lambda : T . X$ as well.

@@

\clearpage
\section{Conversions and Reductions}
\begin{figure}[h]
    \begin{gather*}
        \textbf{Beta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda : T.t)\ u)\ \reduce\ t[u]} 
        \qquad \qquad
        \textbf{Beta-Index} \qquad
        \frac{}{E[\Gamma] \vdash ((\lambda.t)\ u)\ \reduce\ t[u]} 
        \\
        \\
        \textbf{Delta} \qquad
        \frac{\WF(E)[\Gamma] \qquad (c:=t:T) \in E}{E[\Gamma] \vdash c\ \reduce\ t} \\
        \\
        \textbf{Eta-Arrow} \qquad
        \frac{}{E[\Gamma] \vdash \lambda : T . (t\ \$0)\ \reduce\ t}
        \qquad \qquad
        \textbf{Eta-Index} \qquad
        \frac{}{E[\Gamma] \vdash \lambda . (t\ \$0)\ \reduce\ t}
    \end{gather*}
    \caption{Conversions and reductions for the typed lambda calculus.}
\end{figure}

@@

\begin{ruletable}{Rules for sum operators.}
    %
    R-SUM-CONST0
    & $ \sum_{s} 0 \ \reduce\ 0 $ \\
    %
    R-SUM-CONST1
    & $ \sum_{s} \ZEROK(\sigma)\ \reduce\ \ZEROK(\sigma) $ \\
    %
    R-SUM-CONST2
    & $ \sum_{s} \ZEROB(\sigma)\ \reduce\ \ZEROB(\sigma) $ \\
    %
    R-SUM-CONST3
    & $ \sum_{s} \ZEROO(\sigma, \tau)\ \reduce\ \ZEROO(\sigma, \tau) $ \\
    %
    R-SUM-CONST4
    & $ \ONEO(\sigma) \ \reduce\ \sum_{\mathbf{U}(\sigma)} \ket{i} \cdot \bra{i} $
\end{ruletable}

\begin{ruletable}{Rules for eliminating $\delta_{s, t}$. Note that these rules will match the $\delta$ operator modulo the commutativity of its arguments.}
    %
    R-SUM-ELIM0
    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} \delta_{\$i, t} \ \reduce\ \sum_{s_1} \cdots \sum_{s_i}  1$ \\
    %
    R-SUM-ELIM1
    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$i, t} \times \cdots \times a_n) $ \\
    & $ \reduce\ \sum_{s_1} \cdots \sum_{s_i} a_1\{\$i/t\} \times \cdots \times a_n\{\$i/t\} $ \\
    %
    R-SUM-ELIM2
    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (\delta_{\$i, t}.A) \ \reduce\ \sum_{s_1} \cdots \sum_{s_i} A\{\$i/t\} $ \\
    %
    R-SUM-ELIM3
    & $ \$i \text{ free in } t \vdash \sum_{\mathbf{U}(\sigma)} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$i, t} \times \cdots \times a_i).A $ \\
    & $ \reduce\ \sum_{s_1} \cdots \sum_{s_i}  (a_1\{\$i/t\} \times \cdots \times a_n\{\$i/t\}).A\{\$i/t\} $ \\
    R-SUM-ELIM4
    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i}  \delta_{\$(i + 1), \$i} \ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{k_n \in s_i} 1 $ \\
    %
    R-SUM-ELIM5
    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \dots \times \delta_{\$(i+1), \$i} \times \cdots \times a_n) $ \\
    & $ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_n} (a_1\{\$i/\$(i+1)\} \times \cdots \times a_n\{\$i/\$(i+1)\}) $ \\
    %
    R-SUM-ELIM6
    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (\delta_{\$(i+1), \$i}.A) \ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_i} A\{\$i/\$(i+1)\} $ \\
    %
    R-SUM-ELIM7
    & $ \sum_{M} \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1 \times \cdots \times \delta_{\$(i+1), \$i} \times \cdots \times a_n).A $ \\
    & $ \reduce\ \sum_{M} \sum_{s_1} \cdots \sum_{s_i} (a_1\{\$i/\$(i+1)\} \times \cdots \times a_n\{\$i/\$(i+1)\}).A\{\$i/\$(i+1)\} $ \\
\end{ruletable}

\begin{ruletable}{Rules for pushing terms into sum operators}
    %
    R-SUM-PUSH0
    & $ b_1 \times \cdots \times (\sum_{M} a) \times \cdots \times b_n \ \reduce\ \sum_{M} (b_1 \times \cdots \times a \times \cdots \times b_n) $ \\
    %
    R-SUM-PUSH1
    & $ (\sum_{M} a)^* \ \reduce\ \sum_{M} a^* $ \\
    %
    R-SUM-PUSH2
    & $ (\sum_{M} X)^\dagger \ \reduce\ \sum_{M} X^\dagger $ \\
    %
    R-SUM-PUSH3
    & $ a.(\sum_{M} X) \ \reduce\ \sum_{M} (a.X) $ \\
    %
    R-SUM-PUSH4
    & $ (\sum_{M} a).X \ \reduce\ \sum_{M} (a.X) $ \\
    %
    R-SUM-PUSH5
    & $ (\sum_{M} B)\cdot K \ \reduce\ \sum_{M}(B \cdot K) $ \\
    %
    R-SUM-PUSH6
    & $ (\sum_{M} O)\cdot K \ \reduce\ \sum_{M}(O \cdot K) $ \\
    %
    R-SUM-PUSH7
    & $ (\sum_{M} B)\cdot O \ \reduce\ \sum_{M}(B \cdot O) $ \\
    %
    R-SUM-PUSH8
    & $ (\sum_{M} K)\cdot B \ \reduce\ \sum_{M}(K \cdot B) $ \\
    %
    R-SUM-PUSH9
    & $ (\sum_{M} O_1)\cdot O_2 \ \reduce\ \sum_{M}(O_1 \cdot O_2) $ \\
    %
    R-SUM-PUSH10
    & $ B \cdot (\sum_{M} K) \ \reduce\ \sum_{M}(B \cdot K) $ \\
    %
    R-SUM-PUSH11
    & $ O \cdot (\sum_{M} K) \ \reduce\ \sum_{M}(O \cdot K) $ \\
    %
    R-SUM-PUSH12
    & $ B \cdot (\sum_{M} O) \ \reduce\ \sum_{M}(B \cdot O) $ \\
    %
    R-SUM-PUSH13
    & $ K \cdot (\sum_{M} B) \ \reduce\ \sum_{M}(K \cdot B) $ \\
    %
    R-SUM-PUSH14
    & $ O_1 \cdot (\sum_{M} O_2) \ \reduce\ \sum_{M}(O_1 \cdot O_2) $ \\
    %
    R-SUM-PUSH15
    & $ (\sum_{M} X_1) \otimes X_2 \ \reduce\ \sum_{M} (X_1 \otimes X_2) $ \\
    %
    R-SUM-PUSH16
    & $ X_1 \otimes (\sum_{M} X_2) \ \reduce\ \sum_{M} (X_1 \otimes X_2) $
\end{ruletable}
\yx{Note: because we apply type checking on variables, and stick to unique bound variables, these pushing in operations are always sound.}

\begin{ruletable}{Rules for addition and index in sum}
    %
    R-SUM-ADDS0
    & $\sum_{M}(a_1 + \cdots + a_n) \ \reduce\ (\sum_{M} a_1) + \cdots + (\sum_{M} a_n) $ \\
    %
    R-SUM-ADD0
    & $\sum_{M}(X_1 + \cdots + X_n) \ \reduce\ (\sum_{M} X_1) + \cdots + (\sum_{M} X_n) $ \\
    %
    R-SUM-ADD1
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}X + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+1).X $ \\
    %
    R-SUM-ADD2
    & $Y_1 + \cdots + \sum_{i \in M}X + \cdots + \sum_{i \in M}(a.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(1+a).X $ \\
    %
    R-SUM-ADD3
    & $Y_1 + \cdots + \sum_{i \in M}(a.X) + \cdots + \sum_{i \in M}(b.X) + Y_n \ \reduce\ Y_1 + \cdots + Y_n + \sum_{i \in M}(a+b).X $ \\
    %
    R-SUM-INDEX0
    & $ \sum_{\mathbf{U}(\sigma \times \tau)} A \ \reduce\ \sum_{\mathbf{U}(\sigma)} \sum_{\mathbf{U}(\tau)} \textsf{adjust}(A, 1, 2)\{\$0/(\$2, \$1)\} $ \\
    %
    R-SUM-INDEX1
    & $ \sum_{M_1 \star M_2} A \ \reduce\ \sum_{M_1} \sum_{M_2} \textsf{adjust}(A, 1, 2)\{\$0/(\$2, \$1)\} $ \\
    %
    R-SUM-SWAP
    & $ M_1 < M_2 \vdash \sum_{M_2} \sum_{M_1} X \ \reduce\ \sum_{M_1} \sum_{M_2} \textsf{swap01}(X) $
\end{ruletable}
